Hierarchical Dispatching and Overriding

### 2017-06-17 Aplas17/sections/Introduction.tex
Revising overview. Done 2.1

This paper proposes \textit{hierarchical dispatch}: a novel approach
to method dispatching, which combines static and dynamic
information. Using hierarchical dispatch, the method binder will look
at both the \emph{static type} and the \emph{dynamic type} of the
receiver during runtime. When there are multiple branches that cause
unintentional conflicts, the static type can specify one branch among
them for unambiguity, and the dynamic type helps to find the most
specific implementation. In that case both unambiguity and
extensibility are preserved. To present this idea, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. In the model we also propose
\textit{partial overrides}, where method overriding can be applied
only to one branch of the class hierarchy. Our model can be viewed as
a generalization of the trait model by providing additional support to
the triangle inheritance.

In summary, our contributions are: \haoyuan{will have prototype implementation in Scala if we still have time.}
\begin{itemize}
	\item \textbf{Hierarchical dispatch algorithm:} an algorithm that integrates both the static type and dynamic type for method dispatch, and hence
	ensures unambiguity as well as extensibility.
	\item \textbf{Partial overrides:} a novel notion that allows
          methods to override on individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on Featherweight Java, supporting the above features. Some theorems and their proofs
	are attached to confirm the type soundness of the model.\bruno{Anything about unambiguity?}
\end{itemize}

>>>>

This paper proposes \textit{hierarchical dispatch}: a novel approach
to method dispatching, which combines static and dynamic
information. Using hierarchical dispatch, the method binder will look
at both the \emph{static type} and the \emph{dynamic type} of the
receiver during runtime. When there are multiple branches that cause
unintentional conflicts, the static type can specify one branch among
them for unambiguity, and the dynamic type helps to find the most
specific implementation. In that case both unambiguity and
extensibility are preserved. To present this idea, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. In the model we also propose
\textit{hierarchical overriding}, where method overriding can be applied
only to one branch of the class hierarchy. Our model can be viewed as
a generalization of the trait model by providing additional support to
the triangle inheritance.

In summary, our contributions are:
\begin{itemize}
	\item \textbf{Hierarchical dispatch algorithm:} an algorithm that integrates both the static type and dynamic type for method dispatch, and hence
	ensures unambiguity as well as extensibility.
	\item \textbf{Hierarchical overriding:} a novel notion that allows
          methods to override on individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on Featherweight Java, supporting the above features. Some theorems and their proofs
	are attached to confirm the type soundness of the model.\bruno{Anything about unambiguity?}
	\item \textbf{Prototype implementation\footnote{Available in supplementary material attached to the submission.}:} a simple interpreter implemented in Scala that follows our formalization rules, incorporating parsing, type-checking and semantic evaluation.\haoyuan{also mention in formalization.}
\end{itemize}

### 2017-06-17 Aplas17/sections/Overview.tex
UML
Revising overview. Done 2.1

\section{A Running Example: DrawableDeck}~\label{sec:overview}
This section illustrates the features of our \MIM{} model for
resolving unintentional method conflicts. As mentioned before, such a
case arises when two inherited methods happen to have the same
signature, but with different semantics and functionality. This
situation is troublesome to programmers that use multiple
inheritance. Below we illustrate with a running example called
\lstinline|DrawableDeck|, which models a drawable deck of cards. 
Note that we use Java-like syntax
throughout the paper, and all types are defined with the keyword
``\lstinline|interface|''. The concept is closely related to Java 8
interfaces with default methods~\cite{}, and traits. For simplicity, an interface in our model has
the following characteristics:
\begin{itemize}
	\item It allows multiple inheritance.
	\item Every method represents a behaviour and requires a body for its implementation (like Java 8 default methods). There are no abstract methods.
	\item It can be directly instantiated by the \lstinline|new| keyword.
\end{itemize}
Furthermore for simplicity, our formalization does not model state at
this stage. %%, but that should not block our discussions below.
\bruno{I think there are two important things being discussed in this
  section:
1) Hierarchical dispatch: which allows methods calls to be correctly 
dispatched based on a combination of static and dynamic type
information; and
2) Traingle Inheritance: i.e. allowing multiple method implementations 
to coexist in the same class based on the superclass hierarchy.

@@

\subsection{Problem 1: Unintentional Method Conflicts}
Suppose that two components \lstinline|Deck| and \lstinline|Drawable| 
have been developed in a system.  \lstinline|Deck| represents a deck
of cards, and defines \lstinline|draw()| for drawing a card from the
deck.  \lstinline|Drawable| is an interface for graphics that
can be drawn, and also includes a method called \lstinline|draw()| for
visual display. For illustration, the implementation of
\lstinline|draw()| in \lstinline|Drawable| only creates a blank canvas
on the screen, and later some instances (\lstinline|Circle|,
\lstinline|Square|, or others) that extend \lstinline|Drawable| may
define their own \lstinline|drawObject()| methods and invoke
\lstinline|draw()|.

@@

interface Drawable {
  void draw() { // draws something on the screen
    JFrame frame = new JFrame("Canvas");
    frame.setVisible(true);
    ...
  }
}
\end{lstlisting}\vspace{3pt}
\bruno{Space is probably going to be a problem, so you may have to
  consider putting code side-by-side to maximize space usage.}
Note that both methods have \lstinline|void| return type (we will not formalize
\lstinline|void| in our calculus afterwards; here is only for illustration). In \lstinline|Deck|, \lstinline|draw()| tries to get the cards as a stack, pops
out the top card, and so on. While in \lstinline|Drawable|, \lstinline|draw()|
creates a blank canvas using \lstinline|JFrame|. Now, a programmer is designing a
card game with GUI. He may want to draw a deck on the screen, so he defines a drawable
deck using multiple inheritance:

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {...} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is surely for composing the features from various 
components, and to achieve code reuse. Multiple inheritance is supported by many mainstream OO
languages. Nevertheless at this point, \lstinline|DrawableDeck| has to throw a compile
error, since the two \lstinline|draw()| methods cause a conflict, though accidentally.

\subsection{Potential Fixes}

For that problem, there are several workarounds. We go over some of
the main ones next.

\paragraph{I. Delegation.} As an alternative to multiple inheritance,
delegation can be used by introducing two fields with
\lstinline|Drawable| type and \lstinline|Deck| type,
respectively. This avoids method conflicts. Nevertheless, it is known
that using delegation makes it hard to correctly maintain
self-references \haoyuan{?} in an extensible system [], and also
introduces a lot of boilerplate code.

\paragraph{II. Creating a \lstinline|draw()| method in \lstinline|DrawableDeck|, which explicitly overrides the old ones.}
This is a non-solution. It does not make any sense to override both methods with totally different functionalities, as old
methods have to be hidden.

\paragraph{III. Refactor Drawable and/or Deck to rename the methods.} If
the code for \lstinline|Drawable| or \lstinline|Deck| are available
then it may be possible to rename one of the \lstinline|draw|
methods. However this approach is non-modular, as it requires 
modifying existing code, and may not be possible if code is unavailable.

%\paragraph{III. Choosing one of them as the default method, like Mixins.} The mixin model can be applied to choose a
%default one based on linearisation. Similarly, we want to preserve both features, rather than keeping only one of them.

\paragraph{IV. Method exclusion/renaming in traits.} Some trait models
support method exclusion/renaming. Those features
provided can eliminate conflicts, although most
programming languages do not support them. In a traditional OO system,
they can break the subtyping. Moreover, in
contrast with exclusion, renaming can indeed preserve both conflicted
behaviours, however, it is cumbersome in practice, as introducing new
names can affect other code blocks.

\paragraph{V. Static dispatch.}
What we really expect from the language is keeping both methods
without renaming, and the type checker does not complain. 
But we need to find an approach to \emph{disambiguate on method
calls statically}.
\bruno{Perhaps show the solution in MIM here (i.e Section 2.4), and then move to problem
2 (which is being discussed next with shuffleAndDraw)}

Certainly the compiler can ignore the conflict when
\lstinline|DrawableDeck| is declared, but once an object of
\lstinline|DrawableDeck| is created, a method call for
\lstinline|draw()| on that object would be ambiguous. Nonetheless, we can adopt static dispatch for
disambiguating. Some languages like C++ use qualified names in that
way:

\vspace{3pt}\begin{lstlisting}
void func(Drawable obj) {
  obj.draw();
}

DrawableDeck d = new DrawableDeck();
d.Drawable::draw();       // calling draw() in Drawable
((Drawable) d).draw(); // calling draw() in Drawable
func(d);                 // calling draw() in Drawable
\end{lstlisting}\vspace{3pt}
Static dispatch in C++ finds out and invokes the most specific method ``by need''.

>>>>

\section{A Running Example: DrawableDeck}~\label{sec:overview}
This section illustrates the features of our \MIM{} model for
resolving unintentional method conflicts. As mentioned before, such a
case arises when two inherited methods happen to have the same
signature, but with different semantics and functionality. This
situation is troublesome to programmers that use multiple
inheritance. Below we illustrate with a running example called
\lstinline|DrawableDeck|, which models a drawable deck of cards. 
Note that we use Java-like syntax
throughout the paper, and all types are defined with the keyword
``\lstinline|interface|''. The concept is closely related to Java 8
interfaces with default methods~\cite{bono14} and traits. For simplicity, an interface in our model has
the following characteristics:
\begin{itemize}
	\item It allows multiple inheritance.
	\item Every method represents a behaviour and requires a body for its implementation (like Java 8 default methods). There are no abstract methods.
	\item It can be directly instantiated by the \lstinline|new| keyword.
\end{itemize}
Furthermore for simplicity, our formalization does not model state at
this stage. %%, but that should not block our discussions below.
\bruno{I think there are two important things being discussed in this
  section:
1) Hierarchical dispatch: which allows methods calls to be correctly 
dispatched based on a combination of static and dynamic type
information; and
2) Traingle Inheritance: i.e. allowing multiple method implementations 
to coexist in the same class based on the superclass hierarchy.

@@

\subsection{Problem 1: Unintentional Method Conflicts}
Suppose that two components \lstinline|Deck| and \lstinline|Drawable| 
have been developed in a system.  \lstinline|Deck| represents a deck
of cards, and defines \lstinline|draw| for drawing a card from the
deck.  \lstinline|Drawable| is an interface for graphics that
can be drawn, and also includes a method called \lstinline|draw| for
visual display. For illustration, the implementation of
\lstinline|draw| in \lstinline|Drawable| only creates a blank canvas
on the screen, and later some instances (\lstinline|Circle|,
\lstinline|Square|, or others) that extend \lstinline|Drawable| may
define their own \lstinline|drawObject| methods and invoke
\lstinline|draw|.

@@

interface Drawable {
  void draw() { // draws something on the screen
    JFrame frame = new JFrame("Canvas");
    frame.setVisible(true);
    ...
  }
}
\end{lstlisting}\vspace{3pt}
Note that both methods have \lstinline|void| return type (we will not formalize
\lstinline|void| in our calculus afterwards; here is only for illustration). In \lstinline|Deck|, \lstinline|draw| tries to get the cards as a stack, pops
out the top card, and so on. While in \lstinline|Drawable|, \lstinline|draw|
creates a blank canvas using \lstinline|JFrame|. Now, a programmer is designing a
card game with GUI. He may want to draw a deck on the screen, so he defines a drawable
deck using multiple inheritance:

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {...} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is surely for composing the features from various 
components, and to achieve code reuse, as supported by many mainstream OO
languages. Nevertheless at this point, languages like Java simply throw a compile
error on \lstinline|DrawableDeck|, since the two \lstinline|draw| methods cause a conflict, though accidentally.

Now one may quickly comes up with a workaround, which is creating a new \lstinline|draw| in \lstinline|DrawableDeck| to
explicitly overrides the old ones. However, the two conflicting methods have totally different functionalities, merging
them into one does not make any sense. This non-solution will hide the old methods, and break independent extensibility. The
feature we expect is called ``triangle inheritance'', that is to say, when the conflicting methods serve for different behaviours,
we can allow them to coexist in the extending interface in a type-safe way.\\

\noindent\textbf{Potential Fixes} There are several workarounds that come to our mind:
\begin{itemize}
	\item \textbf{I. Delegation.} As an alternative to multiple inheritance,
	delegation can be used by introducing two fields with
	\lstinline|Drawable| type and \lstinline|Deck| type,
	respectively. This avoids method conflicts. Nevertheless, it is known
	that using delegation makes it hard to correctly maintain
	self-references \haoyuan{?} in an extensible system [], and also
	introduces a lot of boilerplate code.
	\item \textbf{II. Refactor Drawable and/or Deck to rename the methods.} If
	the code for \lstinline|Drawable| or \lstinline|Deck| are available
	then it may be possible to rename one of the \lstinline|draw|
	methods. However this approach is non-modular, as it requires 
	modifying existing code, and may not be possible if code is unavailable.
	\item \textbf{III. Method exclusion/renaming in traits.} Some trait models
	support method exclusion/renaming. Those features
	provided can eliminate conflicts, although most
	programming languages do not support them. In a traditional OO system,
	they can break the subtyping. Moreover, in
	contrast with exclusion, renaming can indeed preserve both conflicted
	behaviours, however, it is cumbersome in practice, as introducing new
	names can affect other code blocks.
	\item \textbf{IV. Static dispatch.} Some other languages like C++ have
	direct support for the triangle inheritance. While conflicting methods are
	both preserved, the ambiguity of method call is resolved by static dispatch.
	Qualified names are used to specify the branch to give the right implementation.
\end{itemize}

\noindent\textbf{Solution in \name{}:} below is the pseudo-code that shows the triangle
inheritance supported by \name{}.
We can define \lstinline|DrawableDeck| by inheriting \lstinline|Drawable| and \lstinline|Deck| together while
type safety is ensured. It is slightly different from our prototype as we do not model \lstinline|void|, though one can
define a \lstinline|Void| interface for it.
\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {}

interface Foo {
  void func(Deck obj) { obj.draw(); }
}

// main program
new DrawableDeck().Drawable::draw()      // calling draw() in Drawable
new Foo().func(new DrawableDeck())       // calling draw() in Deck
\end{lstlisting}\vspace{3pt}
Note that both cases in the above code are valid in C++ for disambiguation. As shown by \lstinline|func|, C++ looks at
the static type of the receiver for static dispatching. Normally we can define such a method \lstinline|func| for ``implicit''
casts in \name{}. C++ further allows explicit casts.

In the following text we will see that C++ is still unsatisfactory, as it introduces another problem on code reuse, which motivates
our work.
